---
title: "Harmonization pilot - Invariance testing for depression and anxiety measures in ELSA Brazil - v1"
author: "Pedro Zuccolo, Daniel Fatori, James Lian"
date: "2025-02-18"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: lumen
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
knitr::opts_knit$set(root.dir = '/Users/pedrozuccolo/Desktop/harmonization/data')
output_dir <- "/outputs"
```

# Packages
```{r, echo=TRUE, results="hide", warning=FALSE, message=FALSE}
#MG-CFA - ELSA Brasil
library(lavaan)
library(foreign)
library(haven)
library(tidyverse)
library(lavaanPlot)
library(effectsize)
library(semTools)
library(semPlot)
```

# Depression model 
## Prepare dataset
```{r, echo=TRUE, warning=FALSE, message=FALSE}
#load
elsabr <- read.csv("elsabr_covid.csv") %>%
  select(-X)

dass_items <- elsabr %>%
  #keep depression items for all wave
  select(IDELSA, SEXO, matches("dass13_rec_c*|dass16_rec_c*|dass17_rec_c*|dass21_rec_c*|dass5_rec_c*"))

# definition of items names 
#"mood", "anhe", "guil", "deat", "hope", "fati", "inso", "weig", "apat", "strs", "agit", "cogi","worr", "ganx", "irri", "panc", "impr", "hlth","rest", "phob", "soma", "lone"

#rename dass items
dass <- dass_items %>%
    rename(mood1 = dass13_rec_c1,
           anhe1 = dass16_rec_c1,
           guil1 = dass17_rec_c1,
           hope1 = dass21_rec_c1,
           apat1 = dass5_rec_c1,
           mood2 = dass13_rec_c2,
           anhe2 = dass16_rec_c2,
           guil2 = dass17_rec_c2,
           hope2 = dass21_rec_c2,
           apat2 = dass5_rec_c2,
           mood3 = dass13_rec_c3,
           anhe3 = dass16_rec_c3,
           guil3 = dass17_rec_c3,
           hope3 = dass21_rec_c3,
           apat3 = dass5_rec_c3, 
           mood4 = dass13_rec_c4,
           anhe4 = dass16_rec_c4,
           guil4 = dass17_rec_c4,
           hope4 = dass21_rec_c4,
           apat4 = dass5_rec_c4) 
          
# pivot longer
dass_long <- dass %>%
  pivot_longer(
    cols = matches("^(apat|mood|anhe|guil|hope)[0-9]+$"),
    names_to = c(".value", "Wave"),  
    names_pattern = "([a-z]+)([0-9]+)" 
  ) %>%
  #make wave integer
  mutate(Wave = as.integer(Wave)) %>% # 
  #make dass items binary (simple rule: if dass => 2,then 1, else 0)
  mutate(across(4:last_col(), ~ ifelse(. > 2, 1, 0)))

#convert symptom items to factor
dass.factor <- dass_long %>% mutate(across(4:ncol(dass_long), as.factor))

# Define the list of categorical (ordered) variables
ordered_vars <- c("mood", "anhe", "guil", "hope", "apat")

# Convert them to ordered factors
dass.factor[ordered_vars] <- lapply(dass.factor[ordered_vars], ordered)

str(dass.factor)
```

## check dataset
```{r}
#check if binarization worked
table(dass.factor$mood)
table(dass.factor$anhe)
table(dass.factor$guil)
table(dass.factor$hope)
table(dass.factor$apat) 
```

## Define model
```{r}
# Specify the measurement model
general_model <- '
  # Depression latent variable
  dep =~ mood + anhe + guil + hope + fati + inso + weig + apat + stre + agit + cogn + inde
  
  # Anxiety latent variable
  anx =~ worr + ganx + irri + pani + func + hlth + rest + phob + soma + situ + over + brkd + lone
'

  # Depression latent variable (using Match 1 in symptom item selection spreadsheet: https://docs.google.com/spreadsheets/d/1BjcU4lfUimc4uWSWn4AtmOmctjiTenlI/edit?gid=1202869578#gid=1202869578)

dep_model <- 'dep =~ mood + anhe + guil + hope + apat'
```

## Fit model
### Configural invariance
```{r}
### ------------------------------
### 1. Configural Invariance Model
### ------------------------------
# No equality constraints: the factor structure is the same across groups,
# but loadings and thresholds are freely estimated.
fit_configural <- cfa(dep_model, 
                      data = dass.factor, 
                      group = "Wave", #e.g. 1 is elsa-brasil and 2 is elsa-uk
                      ordered = ordered_vars, 
                      estimator = "WLSMV")

# Display summary for the configural model
summary(fit_configural, standardized = TRUE, modindices = TRUE)
```

### Metric invariance
```{r}
### ------------------------------
### 2. Metric Invariance Model
### ------------------------------
# Constrain factor loadings to be equal across groups.
fit_metric <- cfa(dep_model, 
                  data = dass.factor, 
                  group = "Wave", 
                  ordered = ordered_vars, 
                  estimator = "WLSMV",
                  group.equal = "loadings")

# Display summary for the metric invariance model
summary(fit_metric, standardized = TRUE, modindices = TRUE)
```

### Scalar invariance
```{r}
### ------------------------------
### 3. Scalar Invariance Model
### ------------------------------
# Constrain both factor loadings and thresholds to be equal across groups.
fit_scalar <- cfa(dep_model, 
                  data = dass.factor, 
                  group = "Wave", 
                  ordered = ordered_vars, 
                  estimator = "WLSMV",
                  group.equal = c("loadings", "thresholds"))

# Display summary for the scalar invariance model
summary(fit_scalar, standardized = TRUE, modindices = TRUE)
```

## Fit indices evaluation
* CFI/TLI >.90 (ideally >.95)
* RMSEA <.08 (ideally .06)
* SRMR <.08
* Chi-square p>.05 (doesn't matter if large sample size)

### Configural model
```{r}
fitMeasures(fit_configural, c("cfi", "tli", "rmsea", "srmr"))
```

### Metric model
```{r}
fitMeasures(fit_metric, c("cfi", "tli", "rmsea", "srmr"))
```

### Scalar model
```{r}
fitMeasures(fit_scalar, c("cfi", "tli", "rmsea", "srmr"))
```

# Anxiety model
## Prepare dataset
```{r}
dass_items <- elsabr %>%
  #keep depression items for all wave
  select(IDELSA, SEXO, matches("dass1_rec_c*|dass9_rec_c*|dass14_rec_c*|dass20_rec_c*"))

# definition of items names 
#"mood", "anhe", "guil", "deat", "hope", "fati", "inso", "weig", "apat", "strs", "agit", "cogi","worr", "ganx", "irri", "panc", "impr", "hlth","rest", "phob", "soma", "lone"

#rename dass items
dass <- dass_items %>%
    rename(worr1 = dass9_rec_c1,
           ganx1 = dass20_rec_c1,
           irri1 = dass14_rec_c1,
           rest1 = dass1_rec_c1,
           worr2 = dass9_rec_c2,
           ganx2 = dass20_rec_c2,
           irri2 = dass14_rec_c2,
           rest2 = dass1_rec_c2,
           worr3 = dass9_rec_c3,
           ganx3 = dass20_rec_c3,
           irri3 = dass14_rec_c3,
           rest3 = dass1_rec_c3,
           worr4 = dass9_rec_c4,
           ganx4 = dass20_rec_c4,
           irri4 = dass14_rec_c4,
           rest4 = dass1_rec_c4
           ) 
          
# pivot longer
dass_long <- dass %>%
  pivot_longer(
    cols = matches("^(worr|ganx|irri|rest)[0-9]+$"),
    names_to = c(".value", "Wave"),  
    names_pattern = "([a-z]+)([0-9]+)" 
  ) %>%
  #make wave integer
  mutate(Wave = as.integer(Wave)) %>% # 
  #make dass items binary (simple rule: if dass => 2,then 1, else 0)
  mutate(across(4:last_col(), ~ ifelse(. > 2, 1, 0)))

#convert symptom items to factor
dass.factor <- dass_long %>% mutate(across(4:ncol(dass_long), as.factor))

# Define the list of categorical (ordered) variables
ordered_vars <- c("worr", "ganx", "irri", "rest")

# Convert them to ordered factors
dass.factor[ordered_vars] <- lapply(dass.factor[ordered_vars], ordered)

str(dass.factor)
```

## check dataset
```{r}
#check if binarization worked
table(dass.factor$worr)
table(dass.factor$ganx)
table(dass.factor$irri)
table(dass.factor$rest)
```

## Define model 
```{r}
# Specify the measurement model
general_model <- '
  # Depression latent variable
  dep =~ mood + anhe + guil + hope + fati + inso + weig + apat + stre + agit + cogn + inde
  
  # Anxiety latent variable
  anx =~ worr + ganx + irri + pani + func + hlth + rest + phob + soma + situ + over + brkd + lone
'

  # Depression latent variable (using Match 1 in symptom item selection spreadsheet: https://docs.google.com/spreadsheets/d/1BjcU4lfUimc4uWSWn4AtmOmctjiTenlI/edit?gid=1202869578#gid=1202869578)

anx_model <- 'dep =~ worr + ganx + irri + rest'
```

## Fit model
### Configural invariance
```{r}
### ------------------------------
### 1. Configural Invariance Model
### ------------------------------
# No equality constraints: the factor structure is the same across groups,
# but loadings and thresholds are freely estimated.
fit_configural <- cfa(anx_model, 
                      data = dass.factor, 
                      group = "Wave", #e.g. 1 is elsa-brasil and 2 is elsa-uk
                      ordered = ordered_vars, 
                      estimator = "WLSMV")

# Display summary for the configural model
summary(fit_configural, standardized = TRUE, modindices = TRUE)
```

### Metric invariance
```{r}
### ------------------------------
### 2. Metric Invariance Model
### ------------------------------
# Constrain factor loadings to be equal across groups.
fit_metric <- cfa(anx_model, 
                  data = dass.factor, 
                  group = "Wave", 
                  ordered = ordered_vars, 
                  estimator = "WLSMV",
                  group.equal = "loadings")

# Display summary for the metric invariance model
summary(fit_metric, standardized = TRUE, modindices = TRUE)
```

### Scalar invariance
```{r}
### ------------------------------
### 3. Scalar Invariance Model
### ------------------------------
# Constrain both factor loadings and thresholds to be equal across groups.
fit_scalar <- cfa(anx_model, 
                  data = dass.factor, 
                  group = "Wave", 
                  ordered = ordered_vars, 
                  estimator = "WLSMV",
                  group.equal = c("loadings", "thresholds"))

# Display summary for the scalar invariance model
summary(fit_scalar, standardized = TRUE, modindices = TRUE)
```

## Fit indices evaluation
* CFI/TLI >.90 (ideally >.95)
* RMSEA <.08 (ideally .06)
* SRMR <.08
* Chi-square p>.05 (doesn't matter if large sample size)

### Configural model
```{r}
fitMeasures(fit_configural, c("cfi", "tli", "rmsea", "srmr"))
```

### Metric model
```{r}
fitMeasures(fit_metric, c("cfi", "tli", "rmsea", "srmr"))
```

### Scalar model
```{r}
#fitMeasures(fit_scalar, c("cfi", "tli", "rmsea", "srmr"))
# not extracted because model did not run
```





