---
title: "Harmonization pilot - Invariance testing for depression and anxiety measures in ELSA Brazil - v4"
author: "Pedro Zuccolo, Daniel Fatori, James Lian"
date: "2025-02-18"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: lumen
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
knitr::opts_knit$set(root.dir = '/Users/pedrozuccolo/Desktop/harmonization/data')
output_dir <- "/outputs"
```

# Packages
```{r, echo=TRUE, results="hide", warning=FALSE, message=FALSE}
#MG-CFA - ELSA Brasil
library(lavaan)
library(foreign)
library(haven)
library(tidyverse)
library(lavaanPlot)
library(effectsize)
library(semTools)
library(semPlot)
library(mice)
library(VIM)
```

# Depression model 
## Prepare dataset
```{r, echo=TRUE, warning=FALSE, message=FALSE}
#load
elsabr <- read.csv("/Users/pedrozuccolo/Desktop/harmonization/data/elsabr_covid.csv") %>%
  select(-X)

dass_items <- elsabr %>%
  #keep depression items for all wave
  select(IDELSA, SEXO, matches("dass13_rec_c*|dass16_rec_c*|dass17_rec_c*|dass21_rec_c*|dass5_rec_c*"))

# definition of items names 
#"mood", "anhe", "guil", "deat", "hope", "fati", "inso", "weig", "apat", "strs", "agit", "cogi","worr", "ganx", "irri", "panc", "impr", "hlth","rest", "phob", "soma", "lone"

#rename dass items
dass <- dass_items %>%
    rename(mood1 = dass13_rec_c1,
           anhe1 = dass16_rec_c1,
           guil1 = dass17_rec_c1,
           hope1 = dass21_rec_c1,
           apat1 = dass5_rec_c1,
           mood2 = dass13_rec_c2,
           anhe2 = dass16_rec_c2,
           guil2 = dass17_rec_c2,
           hope2 = dass21_rec_c2,
           apat2 = dass5_rec_c2,
           mood3 = dass13_rec_c3,
           anhe3 = dass16_rec_c3,
           guil3 = dass17_rec_c3,
           hope3 = dass21_rec_c3,
           apat3 = dass5_rec_c3, 
           mood4 = dass13_rec_c4,
           anhe4 = dass16_rec_c4,
           guil4 = dass17_rec_c4,
           hope4 = dass21_rec_c4,
           apat4 = dass5_rec_c4) %>%
  #make dass items binary (simple rule: if dass => 2,then 1, else 0)
  mutate(across(3:last_col(), ~ ifelse(. >= 2, 1, 0)))

# Define the list of categorical (ordered) variables
ordered_vars <- c("mood1", "anhe1", "guil1", "hope1", "apat1", "mood2", "anhe2", "guil2", "hope2", "apat2", "mood3", "anhe3", "guil3", "hope3", "apat3", "mood4", "anhe4", "guil4", "hope4", "apat4")

# Convert them to ordered factors
dass[ordered_vars] <- lapply(dass[ordered_vars], ordered)
```

## check dataset
```{r}
#check if binarization worked
table(dass$mood1)
table(dass$anhe1)
table(dass$guil1)
table(dass$hope1)
table(dass$apat1)
table(dass$mood2)
table(dass$anhe2)
table(dass$guil2)
table(dass$hope2)
table(dass$apat2)
table(dass$mood3)
table(dass$anhe3)
table(dass$guil3)
table(dass$hope3)
table(dass$apat3)
table(dass$mood4)
table(dass$anhe4)
table(dass$guil4)
table(dass$hope4)
table(dass$apat4)
```

## Define general model
```{r}
# Specify the measurement model - simple model
general_model <- '
  # Depression latent variable
  dep =~ mood + anhe + guil + hope + fati + inso + weig + apat + stre + agit + cogn + inde
  
  # Anxiety latent variable
  anx =~ worr + ganx + irri + pani + func + hlth + rest + phob + soma + situ + over + brkd + lone
'
  # Depression latent variable (using Match 1 in symptom item selection spreadsheet: https://docs.google.com/spreadsheets/d/1BjcU4lfUimc4uWSWn4AtmOmctjiTenlI/edit?gid=1202869578#gid=1202869578). This means we are using: mood, anhe, guil, hope, apat
```




## Configural (baseline) invariance
```{r}
### ------------------------------
### 1. Configural Invariance Model
### ------------------------------
# No equality constraints: the factor structure is the same across groups,
# but loadings and thresholds are freely estimated.

## Define configural model

dep_config_model <- '
# Definition of latent factors
dep1 =~ mood1 + anhe1 + guil1 + hope1 + apat1
dep2 =~ mood2 + anhe2 + guil2 + hope2 + apat2 
dep3 =~ mood3 + anhe3 + guil3 + hope3 + apat3
dep4 =~ mood4 + anhe4 + guil4 + hope4 + apat4

# Variances and covariances of latent factors
dep1 ~~ dep1 + dep2 + dep3 + dep4
dep2 ~~ dep2 + dep3 + dep4
dep3 ~~ dep3 + dep4
dep4 ~~ dep4

# Covariances between unique errors over time
mood1 ~~ mood2 + mood3 + mood4
mood2 ~~ mood3 + mood4
mood3 ~~ mood4

anhe1 ~~ anhe2 + anhe3 + anhe4
anhe2 ~~ anhe3 + anhe4
anhe3 ~~ anhe4

guil1 ~~ guil2 + guil3 + guil4
guil2 ~~ guil3 + guil4
guil3 ~~ guil4

hope1 ~~ hope2 + hope3 + hope4
hope2 ~~ hope3 + hope4
hope3 ~~ hope4

apat1 ~~ apat2 + apat3 + apat4
apat2 ~~ apat3 + apat4
apat3 ~~ apat4
'

# Fit
fit_configural <- cfa(dep_config_model, data = dass, 
                         ordered = c("mood1", "anhe1", "guil1", "hope1", "apat1",
                                     "mood2", "anhe2", "guil2", "hope2", "apat2",
                                     "mood3", "anhe3", "guil3", "hope3", "apat3",
                                     "mood4", "anhe4", "guil4", "hope4", "apat4"),
                         parameterization = "theta", estimator = "WLSMV",
                         missing = "pairwise")

```

### Results
```{r}
summary(fit_configural, fit.measures = TRUE, standardized = TRUE)
```

### Visualization
```{r, echo=FALSE}
#add title to semPaths
semPaths(fit_configural, 
  what = "std", 
  whatLabels = "est", 
  layout = "tree2", 
  rotation = 1, 
  edge.label.cex = 0.45, 
  nCharNodes = 10, 
  sizeMan = 3.2,        
  label.cex = 1.2,     
  edge.color = "black"
)
```

### Model indices
* CFI/TLI >.90 (ideally >.95)
* RMSEA <.08 (ideally .06)
* SRMR <.08
* Chi-square p>.05 (doesn't matter if large sample size)

```{r}
fitMeasures(fit_configural, c("cfi", "tli", "rmsea", "srmr"))
```

## Metric invariance
```{r}
### ------------------------------
### 2. Metric Invariance Model
### ------------------------------
# Constrain factor loadings to be equal across groups.

#define metric model
dep_metric_model <- '
  # Constraints factorial loadings to be equal across time points
dep1 =~ V1Loading*mood1 + V2Loading*anhe1 + V3Loading*guil1 + V4Loading*hope1 + V5Loading*apat1

dep2 =~ V1Loading*mood2 + V2Loading*anhe2 + V3Loading*guil2 + V4Loading*hope2 + V5Loading*apat2

dep3 =~ V1Loading*mood3 + V2Loading*anhe3 + V3Loading*guil3 + V4Loading*hope3 + V5Loading*apat3

dep4 =~ V1Loading*mood4 + V2Loading*anhe4 + V3Loading*guil4 + V4Loading*hope4 + V5Loading*apat4


  # Allow covariance between latent factors throughtout time 
  dep1 ~~ dep2 + dep3 + dep4
  dep2 ~~ dep3 + dep4
  dep3 ~~ dep4

  # Allow covariance between errors of the same variables throughout time
  mood1 ~~ mood2 + mood3 + mood4
  anhe1 ~~ anhe2 + anhe3 + anhe4
  guil1 ~~ guil2 + guil3 + guil4
  hope1 ~~ hope2 + hope3 + hope4
  apat1 ~~ apat2 + apat3 + apat4
'
#Fit metric model
fit_metric <- cfa(dep_metric_model, data = dass,
                      ordered = c("mood1", "anhe1", "guil1", "hope1", "apat1",
                                     "mood2", "anhe2", "guil2", "hope2", "apat2",
                                     "mood3", "anhe3", "guil3", "hope3", "apat3",
                                     "mood4", "anhe4", "guil4", "hope4", "apat4"), 
                      parameterization = "theta", estimator = "wlsmv",
                      missing = "pairwise")
```

### Results
```{r}
summary(fit_metric, fit.measures = TRUE, standardized = TRUE)
```

### Visualization
```{r, echo=FALSE}
semPaths(fit_metric, 
         what = "std", 
         whatLabels = "est", 
         layout = "tree2", 
         rotation = 1, 
         edge.label.cex = 0.45, 
         nCharNodes = 10, 
         sizeMan = 3.2,         # size of manifest variables
         label.cex = 1.2,     # font size in manifest variables
         edge.color = "black" # lines color
)
```

### Model indices
* CFI/TLI >.90 (ideally >.95)
* RMSEA <.08 (ideally .06)
* SRMR <.08
* Chi-square p>.05 (doesn't matter if large sample size)
```{r}
fitMeasures(fit_metric, c("cfi", "tli", "rmsea", "srmr"))
```


## Scalar invariance
```{r}
### ------------------------------
### 3. Scalar Invariance Model
### ------------------------------
# Constrain both factor loadings and thresholds to be equal across groups.
#define scalar model

dep_scalar_model <- '
# Constraints factorial loadings to be equal across time points
dep1 =~ V1Loading*mood1 + V2Loading*anhe1 + V3Loading*guil1 + V4Loading*hope1 + V5Loading*apat1

dep2 =~ V1Loading*mood2 + V2Loading*anhe2 + V3Loading*guil2 + V4Loading*hope2 + V5Loading*apat2

dep3 =~ V1Loading*mood3 + V2Loading*anhe3 + V3Loading*guil3 + V4Loading*hope3 + V5Loading*apat3

dep4 =~ V1Loading*mood4 + V2Loading*anhe4 + V3Loading*guil4 + V4Loading*hope4 + V5Loading*apat4

# Allow covariance between latent factors throughtout time 
dep1 ~~ dep2 + dep3 + dep4
dep2 ~~ dep3 + dep4
dep3 ~~ dep4

# Allow covariance between errors of the same variables throughout time
mood1 ~~ mood2 + mood3 + mood4
anhe1 ~~ anhe2 + anhe3 + anhe4
guil1 ~~ guil2 + guil3 + guil4
hope1 ~~ hope2 + hope3 + hope4
apat1 ~~ apat2 + apat3 + apat4


# Constrain thresholds to be equal across time points
mood1 | t_mood*t1
mood2 | t_mood*t1
mood3 | t_mood*t1
mood4 | t_mood*t1

anhe1 | t_anhe*t1
anhe2 | t_anhe*t1
anhe3 | t_anhe*t1
anhe4 | t_anhe*t1

guil1 | t_guil*t1
guil2 | t_guil*t1
guil3 | t_guil*t1
guil4 | t_guil*t1

hope1 | t_hope*t1
hope2 | t_hope*t1
hope3 | t_hope*t1
hope4 | t_hope*t1

apat1 | t_apat*t1
apat2 | t_apat*t1
apat3 | t_apat*t1
apat4 | t_apat*t1
'

# Fit
fit_scalar <- cfa(dep_scalar_model, data = dass,
                      ordered = c("mood1", "anhe1", "guil1", "hope1", "apat1",
                                     "mood2", "anhe2", "guil2", "hope2", "apat2",
                                     "mood3", "anhe3", "guil3", "hope3", "apat3",
                                     "mood4", "anhe4", "guil4", "hope4", "apat4"), 
                      parameterization = "theta", estimator = "wlsmv",
                      missing = "pairwise",
                      control = list(iter.max = 10000,  # Increase max iterations
                               rel.tol = 1e-5))
```


### Results
```{r}
summary(fit_scalar, fit.measures = TRUE, standardized = TRUE)
```

### Visualization
```{r, echo=FALSE}
semPaths(fit_scalar, 
         what = "std", 
         whatLabels = "est", 
         layout = "tree2", 
         rotation = 1, 
         edge.label.cex = 0.45, 
         nCharNodes = 10, 
         sizeMan = 3.2,         # size of manifest variables
         label.cex = 1.2,     # font size in manifest variables
         edge.color = "black" # lines color
)
```

### Model indices
* CFI/TLI >.90 (ideally >.95)
* RMSEA <.08 (ideally .06)
* SRMR <.08
* Chi-square p>.05 (doesn't matter if large sample size)

```{r}
fitMeasures(fit_scalar, c("cfi", "tli", "rmsea", "srmr"))
```


# Anxiety model
## Prepare dataset
```{r}
dass_items <- elsabr %>%
  #keep depression items for all wave
  select(IDELSA, SEXO, matches("dass1_rec_c*|dass9_rec_c*|dass14_rec_c*|dass20_rec_c*"))

# definition of items names 
#"mood", "anhe", "guil", "deat", "hope", "fati", "inso", "weig", "apat", "strs", "agit", "cogi","worr", "ganx", "irri", "panc", "impr", "hlth","rest", "phob", "soma", "lone"

#rename dass items
dass <- dass_items %>%
    rename(worr1 = dass9_rec_c1,
           ganx1 = dass20_rec_c1,
           irri1 = dass14_rec_c1,
           rest1 = dass1_rec_c1,
           worr2 = dass9_rec_c2,
           ganx2 = dass20_rec_c2,
           irri2 = dass14_rec_c2,
           rest2 = dass1_rec_c2,
           worr3 = dass9_rec_c3,
           ganx3 = dass20_rec_c3,
           irri3 = dass14_rec_c3,
           rest3 = dass1_rec_c3,
           worr4 = dass9_rec_c4,
           ganx4 = dass20_rec_c4,
           irri4 = dass14_rec_c4,
           rest4 = dass1_rec_c4
           ) %>% # 
  #make dass items binary (simple rule: if dass => 2,then 1, else 0)
  mutate(across(3:last_col(), ~ ifelse(. > 2, 1, 0)))

# Define the list of categorical (ordered) variables
ordered_vars <- c("worr1", "ganx1", "irri1", "rest1", "worr2", "ganx2", "irri2", "rest2", "worr3", "ganx3", "irri3", "rest3", "worr4", "ganx4", "irri4", "rest4")

# Convert them to ordered factors
dass[ordered_vars] <- lapply(dass[ordered_vars], ordered)
```

## check dataset
```{r}
#check if binarization worked
table(dass$worr1)
table(dass$ganx1)
table(dass$irri1)
table(dass$rest1)
table(dass$worr2)
table(dass$ganx2)
table(dass$irri2)
table(dass$rest2)
table(dass$worr3)
table(dass$ganx3)
table(dass$irri3)
table(dass$rest3)
table(dass$worr4)
table(dass$ganx4)
table(dass$irri4)
table(dass$rest4)
```

## Define general model 
```{r}
# Specify the measurement model
general_model <- '
  # Depression latent variable
  dep =~ mood + anhe + guil + hope + fati + inso + weig + apat + stre + agit + cogn + inde
  
  # Anxiety latent variable
  anx =~ worr + ganx + irri + pani + func + hlth + rest + phob + soma + situ + over + brkd + lone
'

  # Anxiety latent variable (using Match 1 in symptom item selection spreadsheet: https://docs.google.com/spreadsheets/d/1BjcU4lfUimc4uWSWn4AtmOmctjiTenlI/edit?gid=1202869578#gid=1202869578). This means we are using: worr, ganx, irri, rest

anx_model <- 'dep =~ worr + ganx + irri + rest'
```

## Configural invariance
```{r}
### ------------------------------
### 1. Configural Invariance Model
### ------------------------------
# No equality constraints: the factor structure is the same across groups,
# but loadings and thresholds are freely estimated.

#define model
anx_configural_model <- ' 
  # Configural model: No constraints on loadings
  anx1 =~ 1*worr1 + ganx1 + irri1 + rest1
  anx2 =~ 1*worr2 + ganx2 + irri2 + rest2
  anx3 =~ 1*worr3 + ganx3 + irri3 + rest3
  anx4 =~ 1*worr4 + ganx4 + irri4 + rest4

  # Allow covariance between latent factors throughout time 
  anx1 ~~ anx2 + anx3 + anx4
  anx2 ~~ anx3 + anx4
  anx3 ~~ anx4

  # Allow covariance between errors of the same variables throughout time
  worr1 ~~ worr2 + worr3 + worr4
  ganx1 ~~ ganx2 + ganx3 + ganx4
  irri1 ~~ irri2 + irri3 + irri4
  rest1 ~~ rest2 + rest3 + rest4
'

fit_configural <- cfa(anx_configural_model, data = dass,
                      ordered = c("worr1", "ganx1", "irri1", "rest1",
                                  "worr2", "ganx2", "irri2", "rest2",
                                  "worr3", "ganx3", "irri3", "rest3",
                                  "worr4", "ganx4", "irri4", "rest4"), 
                      parameterization = "theta", estimator = "wlsmv",
                      missing = "pairwise")
```

### Results
```{r}
summary(fit_configural, fit.measures = TRUE, standardized = TRUE)
```

### Visualization
```{r, echo=FALSE}
semPaths(fit_configural, 
         what = "std", 
         whatLabels = "est", 
         layout = "tree2", 
         rotation = 1, 
         edge.label.cex = 0.45, 
         nCharNodes = 10, 
         sizeMan = 3.2,         # size of manifest variables
         label.cex = 1.2,     # font size in manifest variables
         edge.color = "black" # lines color
)
```

### Model indices
* CFI/TLI >.90 (ideally >.95)
* RMSEA <.08 (ideally .06)
* SRMR <.08
* Chi-square p>.05 (doesn't matter if large sample size)

```{r}
fitMeasures(fit_configural, c("cfi", "tli", "rmsea", "srmr"))
```


## Metric invariance
```{r}
### ------------------------------
### 2. Metric Invariance Model
### ------------------------------
# Constrain factor loadings to be equal across groups.


#define model
anx_metric_model <- ' 
  anx1 =~ 1*W1Loading*worr1 + W2Loading*ganx1 + W3Loading*irri1 + W4Loading*rest1
  anx2 =~ 1*W1Loading*worr2 + W2Loading*ganx2 + W3Loading*irri2 + W4Loading*rest2
  anx3 =~ 1*W1Loading*worr3 + W2Loading*ganx3 + W3Loading*irri3 + W4Loading*rest3
  anx4 =~ 1*W1Loading*worr4 + W2Loading*ganx4 + W3Loading*irri4 + W4Loading*rest4

  # Allow covariance between latent factors throughout time 
  anx1 ~~ anx2 + anx3 + anx4
  anx2 ~~ anx3 + anx4
  anx3 ~~ anx4

  # Allow covariance between errors of the same variables throughout time
  worr1 ~~ worr2 + worr3 + worr4
  ganx1 ~~ ganx2 + ganx3 + ganx4
  irri1 ~~ irri2 + irri3 + irri4
  rest1 ~~ rest2 + rest3 + rest4
'

fit_metric <- cfa(anx_metric_model, data = dass,
                  ordered = c("worr1", "ganx1", "irri1", "rest1",
                              "worr2", "ganx2", "irri2", "rest2",
                              "worr3", "ganx3", "irri3", "rest3",
                              "worr4", "ganx4", "irri4", "rest4"), 
                  parameterization = "theta", estimator = "wlsmv",
                  missing = "pairwise")
```

### Results
```{r}
summary(fit_metric, fit.measures = TRUE, standardized = TRUE)
```

### Visualization
```{r, echo=FALSE}
semPaths(fit_metric, 
         what = "std", 
         whatLabels = "est", 
         layout = "tree2", 
         rotation = 1, 
         edge.label.cex = 0.45, 
         nCharNodes = 10, 
         sizeMan = 3.2,         # size of manifest variables
         label.cex = 1.2,     # font size in manifest variables
         edge.color = "black" # lines color
)
```

### Model indices
* CFI/TLI >.90 (ideally >.95)
* RMSEA <.08 (ideally .06)
* SRMR <.08
* Chi-square p>.05 (doesn't matter if large sample size)

```{r}
fitMeasures(fit_metric, c("cfi", "tli", "rmsea", "srmr"))
```


## Scalar invariance
```{r}
### ------------------------------
### 3. Scalar Invariance Model
### ------------------------------
# Constrain both factor loadings and thresholds to be equal across groups.

#define model
library(lavaan)

anx_scalar_model <- ' 
  anx1 =~ W1Loading*worr1 + W2Loading*ganx1 + W3Loading*irri1 + W4Loading*rest1
  anx2 =~ W1Loading*worr2 + W2Loading*ganx2 + W3Loading*irri2 + W4Loading*rest2
  anx3 =~ W1Loading*worr3 + W2Loading*ganx3 + W3Loading*irri3 + W4Loading*rest3
  anx4 =~ W1Loading*worr4 + W2Loading*ganx4 + W3Loading*irri4 + W4Loading*rest4

  # Allow covariance between latent factors throughout time 
  anx1 ~~ anx2 + anx3 + anx4
  anx2 ~~ anx3 + anx4
  anx3 ~~ anx4

  # Allow covariance between errors of the same variables throughout time
  worr1 ~~ worr2 + worr3 + worr4
  ganx1 ~~ ganx2 + ganx3 + ganx4
  irri1 ~~ irri2 + irri3 + irri4
  rest1 ~~ rest2 + rest3 + rest4

  # Constraints intercepts to be equal across time points
  worr1 | t_worr*t1
  worr2 | t_worr*t1
  worr3 | t_worr*t1
  worr4 | t_worr*t1

  ganx1 | t_ganx*t1
  ganx2 | t_ganx*t1
  ganx3 | t_ganx*t1
  ganx4 | t_ganx*t1

  irri1 | t_irri*t1
  irri2 | t_irri*t1
  irri3 | t_irri*t1
  irri4 | t_irri*t1

  rest1 | t_rest*t1
  rest2 | t_rest*t1
  rest3 | t_rest*t1
  rest4 | t_rest*t1
'
# fit model
fit_scalar <- cfa(anx_scalar_model, data = dass,
                  ordered = c("worr1", "ganx1", "irri1", "rest1",
                              "worr2", "ganx2", "irri2", "rest2",
                              "worr3", "ganx3", "irri3", "rest3",
                              "worr4", "ganx4", "irri4", "rest4"), 
                  parameterization = "theta", estimator = "wlsmv",
                  missing = "pairwise")
```

### Results
```{r}
summary(fit_scalar, fit.measures = TRUE, standardized = TRUE)
```

### Visualization
```{r, echo=FALSE}
#pdf("anx_scalar_model.pdf", width = 10, height = 7)
semPaths(fit_scalar, 
         what = "std", 
         whatLabels = "est", 
         layout = "tree2", 
         rotation = 1, 
         edge.label.cex = 0.45, 
         nCharNodes = 10, 
         sizeMan = 3.2,         # size of manifest variables
         label.cex = 1.2,     # font size in manifest variables
         edge.color = "black" # lines color
)
#dev.off()
```

### Model indices
* CFI/TLI >.90 (ideally >.95)
* RMSEA <.08 (ideally .06)
* SRMR <.08
* Chi-square p>.05 (doesn't matter if large sample size)

```{r}
fitMeasures(fit_scalar, c("cfi", "tli", "rmsea", "srmr"))
```

